### 2014-10-20 paper/include/background.tex
wrote wrote

The DRAM cell array in each bank is structured as in \reffig{fig:dram_orgb}. One cell consists of a transistor and a capacitor, where the transistor is controlled by the \textit{word line} wire and connects the capacitor to the \textit{bit line} wire. The word line connects a number of cells which forms a row. Each bit line is connected to a sense amplifier, and the row of sense amplifiers constitute the \textit{row buffer}. Data is stored as a charge on the cell capacitor.

@@

The implementation of a DRAM refresh cycle can be made in two different ways, to either affect one row or many. The first is called \textit{RAS-only refresh} (ROR) where a refresh cycle refreshes one targeted row. To target one row, the row's address is put on the bus, and the MC has the responsibility that a row is refreshed within the maximum refresh period. The second is called \textit{CAS before RAS refresh} (CBR). In this scheme, the memory module has an internal address counter that hold which row that will be refreshed. On a refresh cycle, the row stated by the internal address counter is refreshed and the counter is then incremented. Contrary to ROR, CBR do not need provided addresses on the bus, which results in a lower power consumption and less bus usage. % During CBR, a row in each bank is updated, right? if so. write that in the text.

@@

As refresh operations are made by row, the refresh operations will increase linearly with the device capacity. In turn, the throughput loss increases exponentially, as shown in \reffig{fig:dram_throughput}. Additional refresh operations give larger power consumption and the refreshes become a growing portion of the total DRAM energy consumption, as shown in \reffig{fig:dram_power}. When the device capacity reaches \textit{64~GB}, the refresh power nearly constitutes half of the total consumption, an increase from 15\% if the device capacity is 8 GB.

>>>>

The DRAM cell array in each bank is structured as in \reffig{fig:dram_orgb}. One cell consists of a transistor and a capacitor, where the transistor is controlled by the \textit{word line} wire and connects the capacitor to the \textit{bit line} wire. The word line connects a number of cells which forms a row. In turn, one row from each bank form a global row. Each bit line is connected to a sense amplifier, and the row of sense amplifiers constitute the \textit{row buffer}. Data is stored as a charge on the cell capacitor.

@@

The implementation of a DRAM refresh cycle can be made in two different ways, to either affect one row or many. The first is called \textit{RAS-only refresh} (ROR) where a refresh cycle refreshes one targeted row. To target one row, the row's address is put on the bus, and the MC has the responsibility that a row is refreshed within the maximum refresh period. The second is called \textit{CAS before RAS refresh} (CBR). In this scheme, the MC has an internal address counter that hold which global row that will be refreshed. On a refresh cycle, the global row stated by the internal address counter is refreshed and the counter is then incremented. Contrary to ROR, CBR do not need provided addresses on the bus, which results in a lower power consumption and less bus usage. % During CBR, a row in each bank is updated, right? if so. write that in the text.

@@

As refresh operations are made by row, the refresh operations will increase linearly with the device capacity. In turn, the throughput loss increases exponentially, as in \reffig{fig:dram_throughput}. Additional refresh operations give larger power consumption and the refreshes become a growing portion of the total DRAM energy consumption, as in \reffig{fig:dram_power}. When the device capacity reaches \textit{64~GB}, the refresh power nearly constitutes half of the total consumption, an increase from 15\% if the device capacity is 8 GB.

### 2014-10-20 paper/include/flikker.tex
wrote wrote

%flikker
The technique targets Data Tolerance and is proposed by Liu et al. \cite{flikker} and  distinguish data as critical or non-critical. The critical data is refreshed at the normal maximal refresh period, whereas the non-critical data is refreshed at a much lower rate. Refreshing at a lower rate saves refreshing power but increases data corruption. This trade-off was found to be balanced with a refresh period of one second, where the refresh power savings was expected to be $23\%$ and the error rate $4.0 \times 10^{-8}$, if the critical part is $25\%$.

@@

To support the proposed partition scheme, modifications in the operating system, runtime system, and each application are required. In the application, the developer distinguish the program variables as either critical or non-critical. Then, the runtime system uses a custom allocator that places the two types of variables on separate pages. The OS extends the page-table entries one bit, which are utilized by the runtime system to mark the pages depending on its criticality. Pages are then mapped by the OS based on the extended bit, to either the normal rate refreshed part or the low rate refreshed part of a bank. % To detailed.

>>>>

%flikker
The technique is proposed by Liu et al. \cite{flikker} and targets Data Tolerance and Validity. Flikker distinguish data as critical or non-critical. The critical data is refreshed at the normal maximal refresh period, whereas the non-critical data is refreshed at a much lower rate. Refreshing at a lower rate saves refreshing power but increases data corruption. This trade-off was found to be balanced with a refresh period of one second, where the refresh power savings was expected to be $23\%$ and the error rate $4.0 \times 10^{-8}$, if the critical part is $25\%$.

@@

To support the proposed partition scheme, modifications in the operating system, runtime system, and each application are required. In the application, the developer distinguish the program variables as either critical or non-critical. Then the runtime system and OS keep the data of different criticality apart by allocating them on different pages and mapping these pages to the corresponding part of the bank.

### 2014-10-20 paper/include/sparkk.tex
wrote wrote

%Sparkk
Lucas et al. \cite{sparkk} proses Sparkk, a technique based on Data Tolerance and Retention Time. Sparkk gives fine grained control of refresh rates on a banks' rows. Bit mapping is used to distinguish bits with different criticality, where the bits of the same criticality is placed near each other of the global rank row. This scheme gives the advantage to refresh the highly critical bits at one refresh rate, whereas the refresh rate decreases along with the criticality. ADD that it need to have per bank refresh........

The refresh scheme relies on an ordered list of one entry per memory area, which is traversed through by the internal address counter. Each memory area is formed of any number of consecutive rows and can have different refresh periods for each bank. An entry consists of the chosen refresh rates, time-out bits, and the address to the end of the memory area. When the counter address and the address to the end of the memory area matches, the MC advances to the next entry. There, the MC decides if to issue a refresh operation and to which subranks.

\todo[inline]{Drafty below}
Downside: SW modifications. Mapping has to be created for all data types. Somewhat complex -> draws more power = mitigates advantages.

>>>>

%Sparkk
Lucas et al. \cite{sparkk} proses Sparkk, a technique based on Data Tolerance and Validity. Sparkk gives fine grained control of refresh rates on a banks' rows. Bit mapping is used to distinguish bits with different criticality, where the bits of the same criticality is placed near each other in the global row. This scheme gives the advantage to refresh the highly critical bits at one refresh rate, whereas the refresh rate decreases along with the criticality.

The refresh scheme relies on an ordered list of one entry per memory area, which is traversed through by the internal address counter. Each memory area is formed of any number of consecutive rows and can have different refresh periods for each bank. An entry consists of the chosen refresh period for each bank, time-out bits for each bank, and the address to the end of the memory area. When the counter address and the address to the end of the memory area matches, the MC advances to the next entry. There, the MC decides if to issue a refresh operation and if so, which banks to send the command to. The DRAM device has to provide the ability of per bank refresh commands for this to work.

\todo[inline]{Drafty below}

Similar to Flikker, Sparkk needs software support to partition the data. As Sparkk is a more fine grained technique then Flikker, a more extensive support is needed. Lucas et al. proposes a complex .....

Results:
